---
title: 'Blog Post 11: North Carolina Campaign Narrative'
author: John Kulow
date: '2024-12-10'
slug: blog-post-11-north-carolina-campaign-narrative
categories: []
tags: []
---


```{r setup, warning = FALSE, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, messages = FALSE, warning = FALSE)
```

```{r libraries, include=FALSE}
library(blogdown)
library(car)
library(caret)
library(curl)
library(CVXR)
library(foreign)
library(geofacet)
library(ggpubr)
library(ggthemes)
library(glmnet)
library(gridExtra)
library(haven)
library(janitor)
library(kableExtra)
library(knitr)
library(maps)
library(mgcv)
library(mgcViz)
library(mlr3)
library(patchwork)
library(randomForest)
library(ranger)
library(RColorBrewer)
library(readxl)
library(rstan)
library(scales)
library(sf)
library(shinystan)
library(sjPlot)
library(spData)
library(stargazer)
library(tidygeocoder)
library(tidyverse)
library(tigris)
library(tmap)
library(tmaptools)
library(viridis)
library(censable)
library(readstata13)
```


## Introduction
For my campaign narrative analysis, I chose to focus on the state of North Carolina. One of the seven key swing states targeted by the campaigns, North Carolina could have proved crucial to a Harris victory had she been able to flip the state, the only one of the seven that Biden lost in 2020. However, this did not pan out. Donald Trump won the state by 3.2% and with 50.9% of the vote, winning 183,046 more votes than Harris did in the state. This was better than when he took a 49.9% plurality of the vote in 2020, giving him a 1.3% margin and 74,483 vote victory against Biden that year.


## Overview of North Carolina
### Statewide History and Other Notable 2024 Races
Similar to many of its fellow southern states, North Carolina has a fairly conservative electoral history in the modern era. Other than Obama's narrow victory in 2008, North Carolina has not voted for a Democrat in a presidential election since 1976 when southerner Jimmy Carter managed to take the state. 

That said, the state regularly elects Democrats statewide down ballot, even in the Trump era and even with Trump on the ballot. The state has had a Democratic Governor since 2016, with Democratic State Attorney General Josh Stein sweeping to victory this year by almost 15% against scandal-plagued Lieutenant Governor Mark Robinson. Democrats also won a number of other statewide offices this November, including Democratic Congressman Jeff Jackson's 2.9% victory to succeed Josh Stein as Attorney General, extending Democrats' hold on the office that has been uninterrupted since before 1900. Thus, clearly the state is willing to split its ticket, even in today's polarized era. 

Democrats also won the offices of Lt. Governor by 1.9%, Secretary of State by 2.1%, and State Superintendent by 2.3% while losing the races for State Auditor by 1.7%, State Treasurer by 5%,  Agriculture Commissioner by 7.7%, Labor Commissioner by 5.6%, and Insurance Commissioner by 4.3%. One other race that has been - and continues to - make headlines is the race for a State Supreme Court seat that, barring ongoing legal challenges, Democrats are set to win by ~0.013% or 734 votes out of 5,539,609 cast.

In terms of congressional elections, Republicans flipped three seats, however this is only due to a decision by the Republican-controlled State Supreme Court to overturn an anti-gerrymandering ruling from two years prior when the Court was controlled by Democrats, allowing Republicans to mid-cycle redraw the borders of the districts in a heavily partisan way. In 2022, Democrats and Republicans each won 7 districts. In 2024, Democrats easily held onto three districts and narrowly carried the 1st district, the sole competitive district left in the state.

### Internal coalitions
#### Three Main Metro Areas
Despite consistently being a closely contested swing state for much of the 21st century, the state has seen remarkable shifts internally in what its political coalitions look like, and these changes certainly made their mark in the 2024 election as well. The state has a number of unique political regions. First off there are the three largest metro areas in the state: Charlotte, Raleigh-Durham, and Greensboro-High Point-Winston-Salem which is often referred to as the Triad. Each of these cities have majority-Black city centers that historically have voted strongly Democratic. In 2020 and 2024, Republicans have made some gains here, but these areas are still overwhelmingly blue. 

These metro areas each also have heavily populated suburban regions that historically had voted Republican but are now either tied or Democratic-leaning. This shift has mainly happened during the Trump era, although warning signs of slippage for Republicans were present in the Obama era as well. In 2024, even as the state swung nearly 2% to the right, these suburbs remained stagnant and in some areas even shifted left from 2020.

The heavily populated exurban regions also swung left compared to 2020, particularly those in the Charlotte area, with Cabarrus and Lincoln counties sticking out in particular as some of the only counties in the state to swing left this year. 

#### Rural Areas
Although the largest metro areas certainly are the most important politically given their size, rural areas play an outsized role in North Carolinian politics. In fact, North Carolina has the second largest rural population following only Texas. Generally there are three divisions in how the rural areas vote. First off, there are the primarily white eastern and southeastern rural areas. These areas were a key source of strength for the so-called Dixiecrats, but have shifted in the 21st century towards being solidly Republican areas. Although they still sometimes vote Democratic in non-federal races, in the Trump years even down ballot Democrats struggle here now.

Next, there is North Carolina's portion of the so-called Black Belt that stretches across the south. These areas were dominated by plantations before the end of slavery and now host substantial rural Black populations. These counties are concentrated in the Northeast of the state east of Raleigh and some along the border with South Carolina to the southwest of Fayetteville. Historically, these areas had voted overwhelmingly Democratic, with Northampton County on the border with Virginia actually being one of only three counties nationwide to have voted for Democrats in every single Presidential election since 1900. However, with the collapse of Democratic strength with rural whites in eastern North Carolina, Democrats have had to rely entirely on Black voters to win in these areas. In recent years though, poor Black turnout and seemingly accelerating slippage among Democrats with these rural Black voters has allowed Republicans to make significant progress, flipping a number of counties and coming very close to flipping the VRA mandated Black-opportunity 1st congressional district, with Trump managing to win the new gerrymandered version of it by over 3% this year.

Then there's the central and western rural areas. These areas are overwhelmingly white and overwhelmingly conservative, even historically. Without this region, North Carolina would vote solidly blue. Out of just Randolph and Davidson Counties, which are halfway between the Triad and Charlotte, Trump netted about 87,000 votes, or almost half of his margin of victory, winning them by 57% and 47% margins respectively. In 2020, the ~83,000 votes he netted in these two counties was more than his entire ~74,000 margin of victory in the state. Furthermore, these rurals have not seen remarkable shifts in either direction.

#### Appalachian and the Southeast
The final two important regions are quite interesting. First there are the Appalachian mountains in the west. This region leans Republican, with many rural areas that go strongly Republican, albeit by smaller margins than the blood red counties to their east, and with Asheville providing a source of Democratic strength in the region. Asheville, a booming tourist city, has increasingly favored Democrats in recent cycles, with its county shifting leftwards in every election since 2012. This region drew disproportionate media attention because of the impacts of Hurricane Helene, which I will talk about in more depth later.

Then finally, there is the southeast. This region hosts some eastern rural white areas and some Black Belt areas, but I have singled it out because of two mid-sized cities in the region and a notable Native American-plurality county. First there is Fayetteville, which is a heavily non-white city that has shifted more Republican in recent years though remains a Democratic stronghold. Then there is Robeson County, home of the Lumbee Tribe, and which has shifted rapidly Republican in the Trump era, making the Lumbee a rare case of a solidly Republican Native American tribe. Then there are Wilmington and the adjacent Brunswick County. While Wilmington proper votes Democratic, New Hanover County where it sits is split down the middle between the two parties, though it is one of very few counties nationwide that flipped to Democrats in 2020 but that Harris held onto in 2024. Brunswick County is a rapidly growing vacation home area that is dominated by retirement communities, making it a strongly Republican area but one that fits the nationwide trend in 2024 of vacation areas shifting leftwards.

## Forecasts vs Results
### My Prediction
My final state-level prediction had Harris winning North Carolina with 50.3% of the two-party vote. This did not hold up perfectly against reality as she only ended up receiving 48.6%. While only a 1.7% error, this was enough to shift the election towards Trump. This said the lower bound of my model's results did have her down to 45.0% of the two-party vote, meaning that while my model did miss, the final results were within my bounds.

I was also interesting in seeing what my secondary state-level model that I used only for states without public polling showed for North Carolina. That model predicted Harris losing the state with only 47.0% of the vote. Although this would have predicted a Trump victory, it would have been 1.6% off the final result, only 0.1% better than my final prediction.

### Expert Predictions
All major expert forecasts had North Carolina as a Tossup except for Sabato's Crystal Ball, who do not do "Tossup" rankings and ranked North Carolina as a "Lean Republican" state. I believe these forecasts were accurate. North Carolina was a heavily contested state that narrowly went to Trump by only 1.6% more than the nation as a whole. Despite Trump winning it by more than double his 2020 margin in the state, it voted to the left of Biden-won Arizona and only 0.1% to the right of Biden-won Nevada. Perhaps a "Lean Republican" ranking would have been more accurate, but given that many agencies are hesitant to move crucial swing states out from the "Tossup" category, these were reasonable rankings.

## Progression of the Campaign
As one of the seven crucial swing states, North Carolina saw [extensive campaign activity](https://projects.votehub.com/pages/campaign-tracker). Harris help eight public events in the state during the campaign while Trump held fourteen (from the period beginning when Biden dropped out). This included six Harris rallies and eleven Trump rallies, with Harris' first rallies in the state being on September 12th in Charlotte and Greensboro and her final rally being in Charlotte on November 2nd. Trump's first rally was on August 21st in Asheboro and his final rally in the state was in Kinston on November 4th. Both of the major party running mates also campaigned repeatedly in the state. In terms of spending, [Harris outspent Trump](https://www.opensecrets.org/states/NC/presidential/2024) in the state $26,236,424 to $11,993,222, with Trump only outspending her in a few small rural media markets such as Jacksonville and Rocky Mount.

It has been reported repeatedly that behind the scenes a large reason that Democrats put such heavy resources into the state despite their 2020 loss is [Anderson Clayton](https://www.msnbc.com/msnbc-podcast/why-is-this-happening/building-new-south-anderson-clayton-podcast-transcript-rcna170724), the 26 year old state Democratic party chair who ousted the incumbent chair in a leadership election in February 2023 following a disappointing 2022 midterm in the state for Democrats. As state party chair, Clayton brought a renewed energy to state party, consistently outraising the state Republican party, hiring an army of staff members and volunteers, opening a slew of new campaign offices, and rolling out a fresh, more aggressive media strategy. Her focus on local races also meant that North Carolina Democrats fielded candidates in every state senate, congressional, and statewide race and all but two state house districts, which was an incredible candidate recruitment feat. Her tenure as chair so far likely is a key reason why North Carolina did not shift as far to the right as most other states and why Democrats were able to hang on in close races such as the 1st congressional or State Supreme Court seat.

Though, the most notable event of the campaign specific to North Carolina was, of course, Hurricane Helene. Making landfall late on September 26th and hitting North Carolina on the 27th, Hurricane Helene ravaged North Carolina's Appalachian region, a part of the state that rarely sees strong hurricanes. [879,000 North Carolinians lost power](https://carolinapublicpress.org/65628/helene-hammers-nc-with-water-and-wind/) during the storm. [103 North Carolinians died](https://www.ncdhhs.gov/assistance/hurricane-helene-recovery-resources/hurricane-helene-storm-related-fatalities). There were 1,400 landslides. Over 160 water and sewer systems, at least 6,000 miles of road, more than 1,000 bridges and culverts, and roughly 126,000 homes were damaged according to the budget office. There were [at least $53 billion in damages](https://apnews.com/article/north-carolina-hurricane-helene-damage-estimate-1c32acaf158bf17eae815926ea682c0e) in the state. Much of the issue in western North Carolina was flooding, which not only caused significant initial damage but also hampered rescue efforts that were already made difficult by the nature of the mountainous terrain in the region. The disaster was [heavily politicized](https://www.usatoday.com/story/news/politics/elections/2024/09/30/hurricane-helene-damage-trump-georgia/75450820007/), with Republicans heavily criticizing the government's disaster response as being inadequate and delayed. Trump also repeatedly [falsely accused](https://apnews.com/article/kamala-harris-donald-trump-hurricane-helene-3097f1706455929adf5a4eb67f9cdf0f) the federal government and the Democratic Governor of intentionally redirecting aid away from Republican areas. In response, [the Trump campaign requested](https://www.reuters.com/world/us/north-carolina-eases-voting-rules-trump-allies-fret-over-storm-fallout-2024-10-09/) several changes to election rules to allow for better access to absentee ballots in the affected region, as many residents had been displaced and were not there to vote on election day. In terms of the effects on the results, much is unknown. Turnout compared to 2020 seemingly did not shift disproportionately in this region compared to others, though that does not account for differential population changes. 10 counties in the region did shift left compared to 2020, which was very notable given how few counties nationwide and within North Carolina shifted leftwards. 


## Explaining Forecast Error
The first main potential source of error is the issue of the economy. As [Achen and Bartels (2014)](https://press.princeton.edu/books/hardcover/9780691169446/democracy-for-realists?srsltid=AfmBOorl6iN4qgweThT5hVNygA3GYiEx8IZYUksJEgexdWLE3Es2imza), among many others, point out, economic issues are top of mind for voters every cycle. One big narrative emerging after the election is that forecasters alongside the Harris campaign failed to predict just how politically damaging inflation would be for Democrats, the incumbent party. Many forecasters used economic fundamental data like GDP growth that were not abysmal and even occasionally positive for Harris. However, many forecasts including my own did not factor in inflation, the issue that has dominated economic-related headlines throughout the last 2-3 years. Without factoring in inflation, forecasts like mine missed one -if not *the*- key reason behind the nation's rightward shift.

A second potential source of error is the issue of the ground game. Although research by [Enos and Fowler (2016)](https://www-cambridge-org.ezp-prod1.hul.harvard.edu/core/journals/political-science-research-and-methods/article/aggregate-effects-of-largescale-campaigns-on-voter-turnout/20C500B0DE62227873FD24CB3555F779) found that ground campaigns can impact turnout, others such as [Kalla and Broockman (2017)](https://www-cambridge-org.ezp-prod1.hul.harvard.edu/core/journals/american-political-science-review/article/minimal-persuasive-effects-of-campaign-contact-in-general-elections-evidence-from-49-field-experiments/753665A313C4AB433DBF7110299B7433) have found that ground campaigns lack a persuasion effect. Many forecasters took into account Harris' reportedly strong ground game and door-knocking campaign when making their predictions for the swing states. While clearly not enough to change outcomes, there are some signs that Harris' ground game did have some effect that helped forecasts not be as dramatically incorrect as they were for states like New Jersey, Illinois, and New York that rocketed to the right. The swing states saw lower drops in turnout compared to 2020, with four of the six swing states actually having turnout *increases*, making them, alongside semi-competitive New Hampshire, to only states to do so. These swing states also on average shifted to the right notably less than other states. While the ground game does not help to explain why the nation nor these states shifted to the right compared to many forecasts, it does help to explain why these states, in particular North Carolina, had *less* error than many other states did.

## Conclusion
Overall, North Carolina had one of the most interesting and at times chaotic election cycles of any state and even of any swing state. From Hurricanes to cheating scandals and from partisan gerrymandering to hundreds of millions of dollars in ad spending, the state bore the brunt of the idiosyncrasies of modern American politics. In the end, Trump won by larger than predicted, but Democratic advocates for the state's competitiveness won the day as well, managing to keep the margin narrow relative to the nation's swing and securing victories in roughly half of the non-federal statewide elected positions while overperforming the top of the ticket. Democrats continue to struggle with rural voters, and recent trends with non-white voters in the state mean that they will either have to reverse their fortunes with these voters or overperform even more with suburban white voters. With the 2026 North Carolina senate race shaping up to be the top Democratic target alongside Maine on their path to reclaim a senate majority, we will soon see if the Democrats can finally pull together a coalition to win a federal statewide race in the state for the first time since 2008.

```{r, include=FALSE}
## build_site(build_rmd=TRUE)
```