---
title: 'Blog Post 5: Demographics'
author: John Kulow
date: '2024-10-07'
slug: blog-post-5-demographics
categories: []
tags: []
---

```{r setup, warning = FALSE, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, messages = FALSE, warning = FALSE)
```

## Introduction
For this fifth blog post, I want to look at another, likely final, fundamental factor that crucially impacts American elections: demographics. Throughout America's history, demographic voting divisions have impacted elections, from well-observed racial partisan voting gaps to turnout differences in different age groups to the education divide that helped propel Trump's 2016 victory to the increasingly emerging partisan gender gap. Because of these issues and other demographics related-trends, campaigns must keep track of what a state or district's demographics look like. Additionally, although I will not be exploring it in this blog post, demographic *changes* also are noteworthy to campaigns. For example, the fast-growing, rapidly diversifying Atlanta suburbs will be crucial if Harris wants to win in November, whereas slow-growing or even shrinking rural areas in Wisconsin, Michigan, and Pennsylvania pose a key challenge from the Trump campaign who seeks to rack up votes in those heavily Republican areas. As such, it is important to get a sense of what the demographic makeups are of key states in 2024.

## State-Level Demographic Analysis

To begin with, I will look at the breakdown of electorates in terms of race, age, education, and gender in the following 10 key states: Arizona, Florida, Georgia, Michigan, North Carolina, Nevada, Ohio,  Pennsylvania, Texas, and Wisconsin. To do this, I will use a 1% sample of voterfile data provided by Statara Solutions to visualize and analyze these demographic breakdowns.

```{r, include=FALSE}
library(car)
library(caret)
library(CVXR)
library(glmnet)
library(tidyverse)
library(patchwork)
library(knitr)
library(sjPlot)
library(readxl)
library(kableExtra)
library(maps)
library(RColorBrewer)
library(viridis)
library(sf)
library(blogdown)
```

```{r, include=FALSE}
####----------------------------------------------------------#
#### Read, merge, and process data.
####----------------------------------------------------------#

# Read popular vote datasets. 
d_popvote <- read_csv("data/popvote_1948_2020.csv")
d_state_popvote <- read_csv("data/state_popvote_1948_2020.csv")

# Read elector distribution dataset. 
d_ec <- read_csv("data/corrected_ec_1948_2024.csv")

# Read and merge demographics data. 
d_demos <- read_csv("data/demographics.csv")[,-1]

# Read primary turnout data. 
d_turnout <- read_csv("data/turnout_1789_2020.csv")
d_state_turnout <- read_csv("data/state_turnout_1980_2022.csv")
d_state_turnout <- d_state_turnout |> 
  mutate(vep_turnout = as.numeric(str_remove(vep_turnout, "%"))/100) |> 
  select(year, state, vep_turnout)

# Read polling data. 
d_polls <- read_csv("data/national_polls_1968-2024.csv")
d_state_polls <- read_csv("data/state_polls_1968-2024.csv")
```

```{r, include=FALSE}
d_pollav_state <- read_csv("data/state_polls_1968-2024.csv")
```

```{r, include=FALSE}
####----------------------------------------------------------#
#### Voter file sample data
####----------------------------------------------------------#

# Read and merge 1% voterfile data into one dataset. 
voterfile.sample.files <- list.files("data/state_1pc_samples_aug24")

# state level voterfile data
vf_fl <- read_csv("data/state_1pc_samples_aug24/FL_sample.csv") |>
  filter(sii_deceased == 0)
vf_wi <- read_csv("data/state_1pc_samples_aug24/WI_sample.csv") |>
  filter(sii_deceased == 0)
vf_mi <- read_csv("data/state_1pc_samples_aug24/MI_sample.csv") |>
  filter(sii_deceased == 0)
vf_pa <- read_csv("data/state_1pc_samples_aug24/PA_sample.csv") |>
  filter(sii_deceased == 0)
vf_nv <- read_csv("data/state_1pc_samples_aug24/NV_sample.csv") |>
  filter(sii_deceased == 0)
vf_az <- read_csv("data/state_1pc_samples_aug24/AZ_sample.csv") |>
  filter(sii_deceased == 0)
vf_nc <- read_csv("data/state_1pc_samples_aug24/NC_sample.csv") |>
  filter(sii_deceased == 0)
vf_ga <- read_csv("data/state_1pc_samples_aug24/GA_sample.csv") |>
  filter(sii_deceased == 0)
vf_tx <- read_csv("data/state_1pc_samples_aug24/TX_sample.csv") |>
  filter(sii_deceased == 0)
vf_oh <- read_csv("data/state_1pc_samples_aug24/OH_sample.csv") |>
  filter(sii_deceased == 0)
```

```{r, include=FALSE}
rename_demo_data <- function(data) {
  data |>
    mutate(sii_age_range = case_when(
      sii_age >= 18 & sii_age <= 24 ~ "18-24",
      sii_age >= 25 & sii_age <= 34 ~ "25-34",
      sii_age >= 35 & sii_age <= 44 ~ "35-44",
      sii_age >= 45 & sii_age <= 54 ~ "45-54",
      sii_age >= 55 & sii_age <= 64 ~ "55-64",
      sii_age >= 65 ~ "65+",
      TRUE ~ "Unknown"
    )) |>
    mutate(sii_race = case_when(
      sii_race == "W" ~ "White",
      sii_race == "B" ~ "Black",
      sii_race == "H" ~ "Hispanic",
      sii_race == "A" ~ "Asian",
      sii_race == "N" ~ "Native",
      sii_race == "O" ~ "Other",
      TRUE ~ "Unknown"
    )) |>
    mutate(sii_education_level = case_when(
      sii_education_level == "A" ~ "Less than High School",
      sii_education_level == "B" ~ "High School Graduate",
      sii_education_level == "C" ~ "Some College",
      sii_education_level == "D" ~ "Bachelor's Degree",
      sii_education_level == "E" ~ "Graduate Degree",
      TRUE ~ "Unknown"
    ))
}
```

```{r, include=FALSE}
vf_oh <- rename_demo_data(vf_oh)
vf_tx <- rename_demo_data(vf_tx)
vf_ga <- rename_demo_data(vf_ga)
vf_nc <- rename_demo_data(vf_nc)
vf_az <- rename_demo_data(vf_az)
vf_nv <- rename_demo_data(vf_nv)
vf_pa <- rename_demo_data(vf_pa)
vf_mi <- rename_demo_data(vf_mi)
vf_wi <- rename_demo_data(vf_wi)
vf_fl <- rename_demo_data(vf_fl)
```

```{r, include=FALSE}
vf_across_states <- bind_rows(vf_oh, vf_tx, vf_ga, vf_nc, vf_az, vf_nv, vf_pa, vf_mi, vf_wi, vf_fl)
```

```{r}
race_demo <- vf_across_states |>
  mutate(sii_race = if_else(sii_race == "Other" | sii_race == "Unknown", "Other/Unknown", sii_race)) |>
  count(sii_state, sii_race) |>
  group_by(sii_state) |>
  mutate(percentage = n / sum(n) * 100)

## insert race graph
race_demo |>
  mutate(Race = factor(sii_race, levels = c("Native", "Black", "Hispanic", "Asian", "White", "Other/Unknown"))) |>
  ggplot(aes(x = sii_state, y = percentage, fill = Race)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("White" = "#84b0fa",
                               "Black" = "#7bab82",
                               "Hispanic" = "#e4785c",
                               "Asian" = "#fbc345",
                               "Native" = "#eb6bed",
                               "Other/Unknown" = "#9d8b8b")) +
  geom_text(aes(label = paste0(round(percentage, 1), "%")),  
            position = position_stack(vjust = 0.5), 
            color = "black", size = 3) +
  labs(title = "Racial Democraphics in Key States", 
       x = "", 
       y = "Percentage")

```

The above graphic shows the breakdown of various racial groups in these 10 states. Here, we can see the key differences between the midwest and the sun belt. Over 80% of the voterfile sample in the midwestern states of Michigan, Ohio, Pennsylvania, and Wisconsin were white, showing the importance of white voters, and in particular working class white voters, to winning those states. On the other hand you have the sunbelt, where states either have a high Hispanic population, high Black population, or both. 

```{r}
age_demo <- vf_across_states |>
  count(sii_state, sii_age_range) |>
  group_by(sii_state) |>
  mutate(percentage = n / sum(n) * 100)

## insert age graph
age_demo |>
  ggplot(aes(x = sii_state, y = percentage, fill = sii_age_range)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(round(percentage, 1), "%")),  
            position = position_stack(vjust = 0.5), 
            color = "black", size = 3) +
  labs(title = "Age Democraphics in Key States", 
       x = "", 
       y = "Percentage",
       fill = "Age Range")
```

In terms of the breakdown for age groupings, the clear standout is Wisconsin and their significantly larger "unknown" contingent. This is likely due to either an error or record-keeping practices in the state that seemingly leave the ages of younger registered voters severely under-reported. Putting Wisconsin aside, you can notice certain other differences. For instance, Florida, long a retirement hot-spot, boasts the largest 65+ contingent, while fast-growing Texas and Georgia, and to a lesser extent Nevada, have distributions skewed younger compared to the slow-growing, aging populations in midwestern states like Ohio and Pennsylvania.

```{r}
edu_demo <- vf_across_states |>
  count(sii_state, sii_education_level) |>
  group_by(sii_state) |>
  mutate(percentage = n / sum(n) * 100)

## insert edu graph
edu_demo |>
  mutate(edu_level_corrected = factor(sii_education_level, levels = c("Graduate Degree", "Bachelor's Degree", "Some College", "High School Graduate", "Less than High School", "Unknown"))) |>
  ggplot(aes(x = sii_state, y = percentage, fill = edu_level_corrected)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(round(percentage, 1), "%")),  
            position = position_stack(vjust = 0.5), 
            color = "black", size = 3) +
  labs(title = "Education Democraphics in Key States", 
       x = "", 
       y = "Percentage",
       fill = "Education Level")
```

In terms of education numbers, the high, unrepresentative proportion of "less than high school" educated registered voters and the all but insignificant count of registered voters with a bachelor's degree calls into question the accuracy of this data. That said, we can still see some small differences between states, with fairly working class Nevada hosting the largest "less than high school" contingent.

```{r}
gender_demo <- vf_across_states |>
    mutate(sii_gender = case_when(
      sii_gender == "M" ~ "Male",
      sii_gender == "F" ~ "Female",
      TRUE ~ "Other/Unknown"
    )) |>
  count(sii_state, sii_gender) |>
  group_by(sii_state) |>
  mutate(percentage = n / sum(n) * 100)

## insert gender graph
gender_demo |>
  ggplot(aes(x = sii_state, y = percentage, fill = sii_gender)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("Female" = "#fc9ffd",
                               "Male" = "#6a9dfd",
                               "Other/Unknown" = "#bd6afd")) +
  geom_text(aes(label = paste0(round(percentage, 1), "%")),  
            position = position_stack(vjust = 0.5), 
            color = "black", size = 3) +
  labs(title = "Gender Democraphics in Key States", 
       x = "", 
       y = "Percentage",
       fill = "Gender")
```

Finally, in terms of gender we can see that each state has a fairly even distribution of men and women.

## Turnout by Demographic Group by State



```{r}
turnout_percents <- function(df) {
  df |>
    mutate(voted_2020g = ifelse(is.na(svi_vh_2020g), 0, 1)) |>
    group_by(sii_gender, sii_race) |>
    summarise(
      total_registered = n(),
      total_voted = sum(voted_2020g, na.rm = TRUE),
      voting_percentage = (total_voted / total_registered) * 100
    )
}
```

```{r, include=FALSE}
turnout_az <- turnout_percents(vf_az)
turnout_nv <- turnout_percents(vf_nv)
turnout_pa <- turnout_percents(vf_pa)
turnout_wi <- turnout_percents(vf_wi)
turnout_mi <- turnout_percents(vf_mi)
turnout_oh <- turnout_percents(vf_oh)
turnout_tx <- turnout_percents(vf_tx)
turnout_fl <- turnout_percents(vf_fl)
turnout_ga <- turnout_percents(vf_ga)
turnout_nc <- turnout_percents(vf_nc)
```

```{r}
## Turnout Graphs
az_turnout_graph <- turnout_az |>
  filter(sii_race == "White" | sii_race == "Black" | sii_race == "Hispanic" | sii_race == "Asian" | sii_race == "Native") |>
    mutate(sii_gender = case_when(
      sii_gender == "M" ~ "Male",
      sii_gender == "F" ~ "Female",
      sii_gender == "O" ~ "Other",
      TRUE ~ "Unknown"
    )) |>
  filter(sii_gender == "Male" | sii_gender == "Female") |>
  ggplot(aes(x = sii_race, y = voting_percentage, fill = sii_gender)) +
  geom_bar(stat = "identity", 
           position = "dodge") +
  scale_fill_manual(values = c("Female" = "#fc9ffd",
                               "Male" = "#6a9dfd")) +
  ylim(0, 100) +
  labs(title = "2020 Voter Turnout in Arizona",
       x = "Race",
       y = "Turnout %",
       fill = "Gender") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12), 
    legend.position = "none"
  )


fl_turnout_graph <- turnout_fl |>
  filter(sii_race == "White" | sii_race == "Black" | sii_race == "Hispanic" | sii_race == "Asian" | sii_race == "Native") |>
    mutate(sii_gender = case_when(
      sii_gender == "M" ~ "Male",
      sii_gender == "F" ~ "Female",
      sii_gender == "O" ~ "Other",
      TRUE ~ "Unknown"
    )) |>
  filter(sii_gender == "Male" | sii_gender == "Female") |>
  ggplot(aes(x = sii_race, y = voting_percentage, fill = sii_gender)) +
  geom_bar(stat = "identity", 
           position = "dodge") +
  scale_fill_manual(values = c("Female" = "#fc9ffd",
                               "Male" = "#6a9dfd")) +
  ylim(0, 100) +
  labs(title = "2020 Voter Turnout in Florida",
       x = "Race",
       y = "Turnout %",
       fill = "Gender") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12), 
    legend.position = "none"
  )

ga_turnout_graph <- turnout_ga |>
  filter(sii_race == "White" | sii_race == "Black" | sii_race == "Hispanic" | sii_race == "Asian" | sii_race == "Native") |>
    mutate(sii_gender = case_when(
      sii_gender == "M" ~ "Male",
      sii_gender == "F" ~ "Female",
      sii_gender == "O" ~ "Other",
      TRUE ~ "Unknown"
    )) |>
  filter(sii_gender == "Male" | sii_gender == "Female") |>
  ggplot(aes(x = sii_race, y = voting_percentage, fill = sii_gender)) +
  geom_bar(stat = "identity", 
           position = "dodge") +
  scale_fill_manual(values = c("Female" = "#fc9ffd",
                               "Male" = "#6a9dfd")) +
  ylim(0, 100) +
  labs(title = "2020 Voter Turnout in Georgia",
       x = "Race",
       y = "Turnout %",
       fill = "Gender") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12), 
    legend.position = "none"
  )

mi_turnout_graph <- turnout_mi |>
  filter(sii_race == "White" | sii_race == "Black" | sii_race == "Hispanic" | sii_race == "Asian" | sii_race == "Native") |>
    mutate(sii_gender = case_when(
      sii_gender == "M" ~ "Male",
      sii_gender == "F" ~ "Female",
      sii_gender == "O" ~ "Other",
      TRUE ~ "Unknown"
    )) |>
  filter(sii_gender == "Male" | sii_gender == "Female") |>
  ggplot(aes(x = sii_race, y = voting_percentage, fill = sii_gender)) +
  geom_bar(stat = "identity", 
           position = "dodge") +
  scale_fill_manual(values = c("Female" = "#fc9ffd",
                               "Male" = "#6a9dfd")) +
  ylim(0, 100) +
  labs(title = "2020 Voter Turnout in Michigan",
       x = "Race",
       y = "Turnout %",
       fill = "Gender") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12), 
    legend.position = "none"
  )

nc_turnout_graph <- turnout_nc |>
  filter(sii_race == "White" | sii_race == "Black" | sii_race == "Hispanic" | sii_race == "Asian" | sii_race == "Native") |>
    mutate(sii_gender = case_when(
      sii_gender == "M" ~ "Male",
      sii_gender == "F" ~ "Female",
      sii_gender == "O" ~ "Other",
      TRUE ~ "Unknown"
    )) |>
  filter(sii_gender == "Male" | sii_gender == "Female") |>
  ggplot(aes(x = sii_race, y = voting_percentage, fill = sii_gender)) +
  geom_bar(stat = "identity", 
           position = "dodge") +
  scale_fill_manual(values = c("Female" = "#fc9ffd",
                               "Male" = "#6a9dfd")) +
  ylim(0, 100) +
  labs(title = "2020 Voter Turnout in North Carolina",
       x = "Race",
       y = "Turnout %",
       fill = "Gender") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12), 
    legend.position = "none"
  )

nv_turnout_graph <- turnout_nv |>
  filter(sii_race == "White" | sii_race == "Black" | sii_race == "Hispanic" | sii_race == "Asian" | sii_race == "Native") |>
    mutate(sii_gender = case_when(
      sii_gender == "M" ~ "Male",
      sii_gender == "F" ~ "Female",
      sii_gender == "O" ~ "Other",
      TRUE ~ "Unknown"
    )) |>
  filter(sii_gender == "Male" | sii_gender == "Female") |>
  ggplot(aes(x = sii_race, y = voting_percentage, fill = sii_gender)) +
  geom_bar(stat = "identity", 
           position = "dodge") +
  scale_fill_manual(values = c("Female" = "#fc9ffd",
                               "Male" = "#6a9dfd")) +
  ylim(0, 100) +
  labs(title = "2020 Voter Turnout in Nevada",
       x = "Race",
       y = "Turnout %",
       fill = "Gender") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12), 
    legend.position = "none"
  )

oh_turnout_graph <- turnout_oh |>
  filter(sii_race == "White" | sii_race == "Black" | sii_race == "Hispanic" | sii_race == "Asian" | sii_race == "Native") |>
    mutate(sii_gender = case_when(
      sii_gender == "M" ~ "Male",
      sii_gender == "F" ~ "Female",
      sii_gender == "O" ~ "Other",
      TRUE ~ "Unknown"
    )) |>
  filter(sii_gender == "Male" | sii_gender == "Female") |>
  ggplot(aes(x = sii_race, y = voting_percentage, fill = sii_gender)) +
  geom_bar(stat = "identity", 
           position = "dodge") +
  scale_fill_manual(values = c("Female" = "#fc9ffd",
                               "Male" = "#6a9dfd")) +
  ylim(0, 100) +
  labs(title = "2020 Voter Turnout in Ohio",
       x = "Race",
       y = "Turnout %",
       fill = "Gender") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12), 
    legend.position = "none"
  )

pa_turnout_graph <- turnout_pa |>
  filter(sii_race == "White" | sii_race == "Black" | sii_race == "Hispanic" | sii_race == "Asian" | sii_race == "Native") |>
    mutate(sii_gender = case_when(
      sii_gender == "M" ~ "Male",
      sii_gender == "F" ~ "Female",
      sii_gender == "O" ~ "Other",
      TRUE ~ "Unknown"
    )) |>
  filter(sii_gender == "Male" | sii_gender == "Female") |>
  ggplot(aes(x = sii_race, y = voting_percentage, fill = sii_gender)) +
  geom_bar(stat = "identity", 
           position = "dodge") +
  scale_fill_manual(values = c("Female" = "#fc9ffd",
                               "Male" = "#6a9dfd")) +
  ylim(0, 100) +
  labs(title = "2020 Voter Turnout in Pennsylvania",
       x = "Race",
       y = "Turnout %",
       fill = "Gender") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12), 
    legend.position = "none"
  )

tx_turnout_graph <- turnout_tx |>
  filter(sii_race == "White" | sii_race == "Black" | sii_race == "Hispanic" | sii_race == "Asian" | sii_race == "Native") |>
    mutate(sii_gender = case_when(
      sii_gender == "M" ~ "Male",
      sii_gender == "F" ~ "Female",
      sii_gender == "O" ~ "Other",
      TRUE ~ "Unknown"
    )) |>
  filter(sii_gender == "Male" | sii_gender == "Female") |>
  ggplot(aes(x = sii_race, y = voting_percentage, fill = sii_gender)) +
  geom_bar(stat = "identity", 
           position = "dodge") +
  scale_fill_manual(values = c("Female" = "#fc9ffd",
                               "Male" = "#6a9dfd")) +
  ylim(0, 100) +
  labs(title = "2020 Voter Turnout in Texas",
       x = "Race",
       y = "Turnout %",
       fill = "Gender") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12), 
    legend.position = "none"
  )

wi_turnout_graph <- turnout_wi |>
  filter(sii_race == "White" | sii_race == "Black" | sii_race == "Hispanic" | sii_race == "Asian" | sii_race == "Native") |>
    mutate(sii_gender = case_when(
      sii_gender == "M" ~ "Male",
      sii_gender == "F" ~ "Female",
      sii_gender == "O" ~ "Other",
      TRUE ~ "Unknown"
    )) |>
  filter(sii_gender == "Male" | sii_gender == "Female") |>
  ggplot(aes(x = sii_race, y = voting_percentage, fill = sii_gender)) +
  geom_bar(stat = "identity", 
           position = "dodge") +
  scale_fill_manual(values = c("Female" = "#fc9ffd",
                               "Male" = "#6a9dfd",
                               "Unknown" = "#bd6afd")) +
  ylim(0, 100) +
  labs(title = "2020 Voter Turnout in Wisconsin",
       x = "Race",
       y = "Turnout %",
       fill = "Gender") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12), 
    legend.position = "none"
  )
```


``` {r}
az_turnout_graph + fl_turnout_graph
ga_turnout_graph + mi_turnout_graph
nc_turnout_graph + nv_turnout_graph
oh_turnout_graph + pa_turnout_graph
tx_turnout_graph + wi_turnout_graph


## az_turnout_graph + fl_turnout_graph + ga_turnout_graph + mi_turnout_graph + nc_turnout_graph + nc_turnout_graph + oh_turnout_graph + pa_turnout_graph + tx_turnout_graph + wi_turnout_graph
```

*insert analysis here, will do after finishing more code*

## Updated Polling-Based Model

```{r}
d_sept_state <- d_pollav_state |>
  filter(month(poll_date) == 9,
         year > 1996,
         party == "DEM") |>
  group_by(year, state) |>
  select(-candidate) |>
  rename(sept_poll = poll_support)


state_results <- d_state_popvote |>
    select(year, state, D_pv2p)

d_poll_sept_state <- left_join(d_sept_state, state_results, by = c("state", "year")) |>
  drop_na()

reg_sept_state <- lm(D_pv2p ~ sept_poll,
              data = d_poll_sept_state, 
              weights = weeks_left)

tab_model(reg_sept_state, show.se = TRUE, 
          title = "September State-Level Polling Regression Table (2000-2020)", 
          dv.labels = "Democratic Party State-Level Vote Share")
```


```{r}
d_poll_sept_2024_state <- d_pollav_state |>
  filter(month(poll_date) == 9, 
         party == "DEM", 
         year == 2024) |>
  select(state, weeks_left, poll_date, poll_support) |>
  rename(sept_poll = poll_support)

pred_2024_state <- predict(reg_sept_state, d_poll_sept_2024_state, interval = "prediction")

prediction_state <- pred_2024_state[, "fit"]
lower_state <- pred_2024_state[, "lwr"]
upper_state <- pred_2024_state[, "upr"]

sept_2024_table_state <- d_poll_sept_2024_state |>
  mutate(predicted_values = prediction_state,
         lower_bounds = lower_state,
         upper_bounds = upper_state) |>
  group_by(state) |>
  summarize(mean_prediction = round(weighted.mean(predicted_values, weeks_left, na.rm = TRUE),2), 
            mean_lower_bound = round(weighted.mean(lower_bounds, weeks_left, na.rm = TRUE),2), 
            mean_upper_bound = round(weighted.mean(upper_bounds, weeks_left, na.rm = TRUE),2)
            )
```


```{r}
knitr::kable(sept_2024_table_state, 
             col.names = c("State", "Harris Vote Share", "Lower Bound", "Upper Bound"))
```






















```{r, include=FALSE}
## build_site(build_rmd=TRUE)
```
