---
title: 'Blog Post 1: Exploring Past Presidential Election Results'
author: John Kulow
date: 'September 9th, 2024'
slug: blog-post-1-exploring-past-presidential-election-results
categories: []
tags: []
---

## **Introduction**

In this first blog post, I seek to gain a better understanding of past U.S. presidential elections that have taken place after World War Two. This analysis, which will include an exploration of national popular vote margins, state-level margins, and state-level swings, will help guide my first predictive model for the 2024 presidential election.

## **Historical Two-Party Vote Share**
### National Results:

insert national 2pv paragraph here

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(maps)
library(tidyverse)
```

```{r}
d_popvote <- read_csv("data/popvote_1948-2020.csv")

d_popvote_wide <- d_popvote |>
    select(year, party, pv2p) |>
    pivot_wider(names_from = party, values_from = pv2p)
d_popvote_wide <- d_popvote_wide |> 
    mutate(winner = case_when(democrat > republican ~ "D",
                              TRUE ~ "R"))

```

```{r}
d_popvote |>
  ggplot(aes(x = year, y = pv2p, color = party)) +
  geom_line() + 
  scale_color_manual(values = c("dodgerblue4", "firebrick1")) +
  theme_bw()
```


### State-Level Results:

insert state winner and state margin paragraphs here

### State-Level Swings:

insert state swing paragraph here

## **Initial Predictions**
### Prediction 1: Based off of Past Election Results

Prediction one paragraph here

### Prediction 2: Adding in OLS

OLS prediction 2 paragraph here









